---
output:
  html_document: default
  pdf_document: default
---
WHITE WINE ANALYSIS by MANUEL 
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE}

setwd('/Users/manuelfreude/AnacondaProjects/03 R')

local({r <- getOption("repos")
       r["CRAN"] <- "http://cran.r-project.org" 
       options(repos=r)})
```

```{r include=FALSE}

install.packages("ggplot2", dependencies = T, 
                 repos = "http://cran.us.r-project.org") 
install.packages("knitr", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("dplyr", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("gridExtra", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("corrplot", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("GGally", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("psych", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("reshape", dependencies = T,
                 repos = "http://cran.us.r-project.org")
install.packages("memisc", dependencies = T)
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(corrplot)
library(GGally)
library(psych)
library(reshape)
library(memisc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load the Data
ww <- read.csv('wineQualityWhites.csv')
head(ww)
```

There are 4898 white wines in our dataset with 12 features for EDA.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(ww)
```

# Univariate Commented Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = quality, data = ww, fill = I("brown")) +
  scale_x_continuous(breaks = c(3, 4, 5, 6, 7, 8, 9)) +
  ggtitle('White wine quality')

summary(ww$quality)

```

The quality looks rather normally distributed, no transformation needed. 

We can see here that the majority of white whines has a quality rating between
5 and 7, there are very limited counts at the outer boundaries of the histogram.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = fixed.acidity, data = ww, binwidth = 0.1, fill = I("brown")) +
  scale_x_continuous(breaks = c(4, 5, 6, 7, 8, 9, 10, 11), limits = c(4, 11)) +
  ggtitle('White wine fixed acidity')

```

The fixed acidity looks rather normally distributed, maybe a little skewed to 
the right. I observed some outliers below 4 or beyond 11, which are not included 
in the histogram. I reduced the binwidth for a better view. For now, we keep
the skewness in mind as it might be worth testing log10 transformation later on.


```{r echo=FALSE, message=FALSE, warning=FALSE}

p1 <- qplot(x = volatile.acidity, data = ww, binwidth = 0.01, 
            fill = I("brown")) +
  scale_x_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7), 
                     limits = c(0.1, 0.7)) +
  ggtitle('White wine volatile acidity')


p2 <- qplot(x = log10(volatile.acidity), data = ww, binwidth = 0.01, 
            fill = I("orange")) +
  ggtitle('White wine volatile acidity (log10)')

grid.arrange(p1, p2, ncol = 1)
  
```

Looking at the histogram for the volatile acidity, this appears even more right
skewed than the fixed acidity. I set the histogram limits from 0.1 to 0.7, 
knowing that there are wines with more volatile acidity in our population. The 
second chart shows the log10 transformed version, which comes much closer to a 
normal distribution, we might be able to use that later on.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = citric.acid, data = ww, binwidth = 0.01, fill = I("brown")) +
  scale_x_continuous(breaks = c(0.0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8), 
                     limits = c(0.0, 0.8)) +
  ggtitle('White wine citric acid')

```

Citric acid looks rather normally distributed, even though we have surprisingly
many wines at 0.48/0.49 and 0.73/0.74. 


```{r echo=FALSE, message=FALSE, warning=FALSE}

p3 <- qplot(x = residual.sugar, data = ww, fill = I("orange"), binwidth = 0.5) +
  scale_x_continuous(breaks = seq(from = 0, to = 45, by = 5)) +
  ggtitle('White wine residual sugar')

p4 <- ggplot(aes(x = 1, y = residual.sugar), data = ww, 
             fill = I("brown")) +
  geom_jitter(alpha = 0.1) +
  geom_boxplot(alpha = 0.2, color = 'yellow') +
  ggtitle('White wine residual sugar outliers')

p5 <- qplot(x = log10(residual.sugar), data = ww, fill = I("orange")) +
  ggtitle('White wine residual sugar (log10)')

grid.arrange(p3, p4, p5, ncol = 1)

```

Residual sugar is clearly skewed to the right, many wines are at a peak below 
2.5. The boxplot over the scattered points shows the high number of outliers in
the fourth quartile. The log10 doesn't look like a bell curve, there is a drop 
in the middle, let's see how we can deal with this later on.

```{r echo=FALSE, message=FALSE, warning=FALSE}

p6 <- qplot(x = chlorides, data = ww, binwidth = 0.001, fill = I("brown")) +
  ggtitle('White wine chlorides')

p7 <- qplot(x = log10(chlorides), data = ww, binwidth = 0.01, 
            fill = I("orange")) +
  ggtitle('White wine chlorides (log10)')

grid.arrange(p6, p7, ncol = 1)

```

Chlorides are a little right-skewed, the log10 transformation puts it in a 
rather normal distribution.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = free.sulfur.dioxide, data = ww, binwidth = 1, fill = I("brown")) +
  xlim(0, 80) +
  ggtitle('White wine free sulfur dioxide')

```

Free sulfur dioxide looks rather normally distributed. Again, I set the xlims.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = total.sulfur.dioxide, data = ww, binwidth = 5, fill = I("brown")) +
  xlim(0, 275) +
  ggtitle('White wine total sulfur dioxide')

```

Total sulfur dioxide also looks rather normally distributed. Limits for x were 
modified.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = density, data = ww, binwidth = 0.00005, fill = I("brown")) +
  xlim(0.9875, 1.005) +
  ggtitle('White wine density')

```

Density is interesting, distribution rather normal but it seems that there are
high counts followed by low counts followed by high counts when we move along
the x axis. Maybe something due to measuring with different lab equipment.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = pH, data = ww, binwidth = 0.01, fill = I("brown")) +
  ggtitle('White wine pH')

```

pH value looks straightforward normally distributed. No xlimits this time.

```{r echo=FALSE, message=FALSE, warning=FALSE}

p8 <- qplot(x = sulphates, data = ww, binwidth = 0.01, fill = I("brown")) +
  ggtitle('White wine sulphates')

p9 <- qplot(x = log10(sulphates), data = ww, binwidth = 0.005, 
            fill = I("orange")) +
  ggtitle('White wine sulphates (log10)')

grid.arrange(p8, p9, ncol = 1)

```

Sulphates are again right-skewed, the log10 transformation creates our normal 
distribution. No xlimites this time.


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(x = (alcohol), data = ww, binwidth = 0.1, fill = I("brown")) +
  ggtitle('White wine alcohol')

```

Alcohol is to some extent a special case. It's not really normally distributed, 
but transformations with log10 or sqrt don't change that, might be more depth of
transformations needed at a later stage.


# Univariate Analysis

### What is the structure of your dataset?

There are 4898 white wines in our dataset with 12 numerical features: 

Input variables (based on physicochemical tests):
  1 - fixed acidity (tartaric acid - g / dm^3)
  2 - volatile acidity (acetic acid - g / dm^3)
  3 - citric acid (g / dm^3)
  4 - residual sugar (g / dm^3)
  5 - chlorides (sodium chloride - g / dm^3
  6 - free sulfur dioxide (mg / dm^3)
  7 - total sulfur dioxide (mg / dm^3)
  8 - density (g / cm^3)
  9 - pH
  10 - sulphates (potassium sulphate - g / dm3)
  11 - alcohol (% by volume)

Output variable (based on sensory data): 
  12 - quality (score between 0 and 10)

Key statistics for these variables are listed below: 

  1 - fixed acidity (range: 3.80 - 14.20, mean: 6.86)
  2 - volatile acidity (range: 0.08 - 1.10, mean: 0.28)
  3 - citric acid (range: 0.00 - 1.66, mean: 0.33)
  4 - residual sugar (range: 0.60 - 65.80, mean: 6.39)
  5 - chlorides (range: 0.01 - 0.35, mean: 0.05)
  6 - free sulfur dioxide (range: 2.00 - 289.00, mean: 35.31)
  7 - total sulfur dioxide (range: 9.00 - 440.00, mean: 138.40)
  8 - density (range: 0.99 - 1.04, mean: 0.99)
  9 - pH (range: 2.72 - 3.82, mean: 3.19)
  10 - sulphates (range: 0.22 - 1.08, mean: 0.49)
  11 - alcohol (range: 8.00 - 14.20, mean: 10.51)
  12 - quality (range: 3.00 - 9.00, mean: 5.88)


Details for quality, our dependent variable aka the main feature of interest in 
our dataset for my analysis: 

 Min. 1st Qu.   Median  Mean    3rd Qu.  Max. 
3.000   5.000   6.000   5.878   6.000   9.000 

For exploring what contributes to the quality derived from sensory tests, I will
have a closer look at all other variables, maybe some new learnings are in for 
me as I'm not the white wine expert beyond knowing what alcohol and pH mean.


Other observations:

We can see here that the majority of white whines has a quality rating between
5 and 7, there are very limited counts at the outer boundaries.

Binwidth modifications and xlimit reductions were usually helpful for the plots,
many variables are more or less skewed, straining the plot limits.

I made some log10 transformations to get a better plot with regards to the
normal distribution, specifically for volatile acidity, residual sugar, 
chlorides and sulphates. This information will be useful for later analysis as
we should try to work with normal distributions to avoid misleading results.

Residual sugar is clearly skewed to the right, many wines are at a peak below 
2.5. The log10 doesn't look like a bell curve, there is a drop in the middle, 
let's see how we can deal with this later on.

Density is interesting, distribution rather normal but it seems that there are
high counts followed by low counts followed by high counts when we move along
the x axis. Maybe something due to measuring.

Alcohol is to some extent a special case. It's not really normally distributed, 
but transformations with log10 or sqrt don't change that, might be more depth of
transformations needed at a later stage, maybe we can also work with ratios.

#### Sources: 

https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt

White Wines dataset



# Bivariate Commented Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

cor_mat <- cor(ww)
corrplot(cor_mat, method = "circle", title = "Correlation matrix", 
         mar=c(0,0,1,0)) 

```

That's helpful. We're focussing on quality, so here are some observations:

 - lower volatile acidity seems to have a slightly positive influence
 - same applies for chlorides and total sulfur dioxide
 - lower density seems to have a rather positive influence on quality
 - higher alcohol seems to have a strongly positive impact on quality 

Amongst the others, following observations are interesting: 

 - density has a strong negative correlation with alcohol and a strong positive
   relation with residual sugar, alcohol has a strong negative correlation with
   residual sugar. These are the strongest effects observed 
 - pH value is negatively related to fixed acidity
 - levels of free sulfur dioxide and total sulfur dioxide are positively 
   correlated with each other and positively correlated with density
 - fixed acidity is not related to volatile acidity
 - chlorides are positively correlated with density, but negatively with alcohol

We need to keep in mind the rule of thumb of small relatedness starting with 
r >= 0.3, which we can show here for some cases. Yet, this also depends on the 
case number, we have plenty of cases in our dataset and will test significance
in regression models. 

With the univariate histograms and correlation plot we're good to move on with 
more bivariate plots.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ww$factor.quality <- factor(ww$quality)
ggplot(aes(factor.quality, volatile.acidity), data = ww) + 
  geom_jitter(alpha = 0.4) +
  geom_boxplot(color = 'orange') + 
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  ggtitle('Quality vs. volatile acidity')

```
```{r echo=FALSE, message=FALSE, warning=FALSE}

describeBy(ww$volatile.acidity, group = ww$quality, mat=T)

```

The boxplot underlines a slightly negative correlation, I factored quality to
get the desired view. Looking at the means ('x'), the highest appear at quality 
levels of 3 & 4. Here, the third quartiles are spread out quite a bit. We can 
also see a rather large number of outliers of volatile acidity at quality 
levels 5 & 6.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(quality, chlorides), data = ww) + 
  geom_jitter(alpha = 1/5, position = position_jitter(h=0)) +
  scale_x_continuous(breaks = 2:10) +
  geom_smooth(method = "lm", se = FALSE,size=1, color = 'orange') +
  ylim (quantile(ww$chlorides, 0), quantile(ww$chlorides, 0.95)) +
  ggtitle('Quality vs. chlorides (chlorides outliers removed)')

```

For chlorides, the correlation coefficient wiht quality was slighly negative. 
This can be confirmed by looking at the plot, medium chlorides are centered at
medium quality (level 6), higher chlorides at quality level 5 and lower 
chlorides at quality level 7.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(factor.quality, total.sulfur.dioxide), data = ww) + 
  geom_jitter(alpha = 0.4) +
  geom_boxplot(color = 'orange') + 
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  ggtitle('Quality vs. total sulfur dioxide')

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

describeBy(ww$total.sulfur.dioxide, group = ww$quality, mat=T)

```

The boxplot using factored quality and total sulfur dioxide shows a sligthly 
negative relation. However, quality level 3 shows some outliers to high total
sulfur dioxide levels. Removing these shows a rather leveled picture, so that
the relation between quality and total sulfur dioxide does not show a clear
direction.



```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(quality, density), data = ww) + 
  geom_jitter(alpha = 1/5, position = position_jitter(h=0), 
              color = 'orange') +
  scale_x_continuous(breaks = 2:10) +
  ylim (quantile(ww$density, 0), quantile(ww$density, 0.95)) + 
  geom_line(stat = 'summary', fun.y = mean) +
  ggtitle('Quality vs. density (density outliers removed)')

```

The scatterplot of quality and density confirms the negative relation. The line
shows the density mean per quality, and is going towards lower density values, 
the higher the wine quality becomes. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(quality, alcohol), data = ww) + 
  geom_point(alpha = 1/5, color = 'orange', position = 'jitter') +
  scale_x_continuous(breaks = 2:10) +
  geom_line(stat = 'summary', fun.y = mean) +
  ggtitle('Quality vs. alcohol')

```

The scatterplot of quality and alcohol confirms the positive relation. The line
shows the alcohol mean per quality, and is going towards higher alcohol values, 
the higher the wine quality becomes. We need to watch out as we have most data
points at quality levels 5-7, so should not draw the conclusion that extremely
high alcohol levels will automatically lead to better wine quality ratings.

Interesting point is that density and alcohol have opposite effects on quality.
Both are related to residual sugar, we will have a look at that.

```{r echo=FALSE, message=FALSE, warning=FALSE}

p10 <- ggplot(aes(density, alcohol), data = ww) + 
  geom_point(color = 'orange') +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple') +
  ggtitle('Density vs. alcohol (density outliers removed)')

p11 <- ggplot(aes(density, residual.sugar), data = ww) + 
  geom_point(color = 'orange') +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) +
  ylim (quantile(ww$residual.sugar, 0), quantile(ww$residual.sugar, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple') +
  ggtitle('Density vs. residual sugar (density/sugar outliers removed)')

p12 <- ggplot(aes(alcohol, residual.sugar), data = ww) + 
  geom_point(color = 'orange') +
  ylim (quantile(ww$residual.sugar, 0), quantile(ww$residual.sugar, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple') +
  ggtitle('Alcohol vs. residual sugar (residual sugar outliers removed)')

grid.arrange(p10, p11, p12, ncol = 1)

```

Here, our suspicion of a relation between some of our independent variables is 
confirmed. We can observe the following based on scatter plots with a linear 
regression smoother (purple): 

 - Density increases with lower alcohol levels
 - Density increases with higher residual sugar levels
 - Alcohol decreases with higher residual sugar levels

The smoother lines confirm the correlation coefficients. This example also 
confirms that if we want to run a stellar regression analysis of what impacts
white wine quality, we need to be aware that the independent variables are 
related with one another. Without me being the wine expert, it might be the 
case that the level of residual sugar in a wine influences both density and 
alcohol and therefore indirectly the quality, even though sugar does not show
highly significant relations to wine quality in our correlation matrix.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(pH, fixed.acidity), data = ww) + 
  geom_point(color = 'orange', position = 'jitter') +
  ylim (quantile(ww$fixed.acidity, 0), quantile(ww$fixed.acidity, 0.95)) +
  geom_line(stat = 'summary', fun.y = mean) +
  ggtitle('pH vs. fixed acidity (fixed acidity outliers removed)')

```

One last scatterplot allows the conclusion that pH values decrease when fixed
acidity increases. The mean line especially between pH 3.0 - 3.4 goes rather
smooth, shows some volatility below or beyond that interval.


# Bivariate Analysis Summary

### Box-, Scatter-, Line- and Smootherplots

We could confirm most selected correlations with the plots as described below
each plot. In one case, quality vs. total sulfur dioxide, the plot incl. 
outliers confirms the correlation, but would probably not do so when removing
the outliers. This showcases that the correlation matrix should not be viewed
as the single source of truth for further analysis.

### Multicollinearity

The observations between density, alcohol and residual sugar indicate that some
of the independent variables are interlinked. One could look at some initial 
R-squared values between quality and other variables here. But to avoid
misleading conclusions, I will do that in the multivariate section rather than
here in the bivariate analysis.

### Next steps

It will be interesting to enrich some of the plots with a third variable in the
next section. Also, overall regression models will be developed to gain a high
level picture, yet not running many model fit checks, e.g. against multicolli-
nearity at we can do with Python VIF.

#### Sources: 

https://stackoverflow.com/questions/40509217/how-to-have-r-corrplot-title-position-correct

https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html


# Multivariate Commented Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(alcohol, density, color = factor(quality)), data = ww) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE, size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  ylim (quantile(ww$density, 0), quantile(ww$density, 0.95)) + 
  ggtitle('Alcohol vs. density (density outliers removed)')

```

A closer look at the quality vs. density reveals our hypothesis. The lines by 
factored quality are all indicating negative relations, this observation lets us
conclude that the level of alcohol declines with increasing density. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, alcohol), data = ww) + 
  geom_point(aes(color = factor.quality)) +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) + 
  ggtitle('Quality vs. density (density outliers removed)')

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, alcohol), data = ww) + 
  geom_point(aes(color = factor.quality)) +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) + 
  facet_wrap(~factor.quality) +
  geom_smooth(method = "lm", se = FALSE, size=1)  +
  ggtitle('Quality vs. density facet wrap (density outliers removed)')

```

The two plots above visualize the negative relation between density and alcohol, 
basically no matter which quality we are looking at. The facet wrapper shows 
that the relation is valid for many quality levels, the highest level 9 only
has a small number of data points.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(density, residual.sugar, color = alcohol), data = ww) + 
  geom_point(position = 'jitter') +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) +
  ylim (quantile(ww$residual.sugar, 0), quantile(ww$residual.sugar, 0.95)) +
  ggtitle('Density vs. residual sugar (density/sugar outliers removed)')

```

Our multicollinear relation between residual sugar, density and alcohol is indi-
cated here. We can see the positive relation between residual sugar and density, 
yet it appears that the alcohol level decreases with increasing density. 

Interesting is the observation that the left side of the plot is rather light 
blue, meaning that also wines with high residual sugar might have high alcohol
levels. We can clearly see here that density is more negatively related to 
alcohol than residual sugar is to alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(pH, fixed.acidity, color = factor.quality), data = ww) + 
  geom_point(position = 'jitter') +
  ylim (quantile(ww$fixed.acidity, 0), quantile(ww$fixed.acidity, 0.95)) +
  ggtitle('pH vs. fixed acidity (fixed acidity outliers removed)')

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(pH, fixed.acidity), data = ww) + 
  geom_point(aes(color = factor.quality), position = 'jitter') +
  ylim (quantile(ww$fixed.acidity, 0), quantile(ww$fixed.acidity, 0.95)) +
  facet_wrap (~factor.quality) +
  geom_smooth(method = "lm", se = FALSE, size=1)  +
  ggtitle('pH vs. fixed acidity facet wrap (fixed acidity outliers removed)')

```

Here is a nice representation that shows the relation of pH level and fixed 
acidity is not influencing the quality level.  

# Multivariate Plot Analysis Summary

Our plots confirmed the main results from the bivariate section. Now let's look
in more depth into the data by building regression models. For these, I will 
first run a model with the plain variables "as they are" and then come back to 
the transformations that were introduced in the univariate chapter to use
bell-curve like distributions.

# Multivariate Regression Commented Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE}

m1 <- lm(I(quality) ~ density, data = ww)
m2 <- update(m1, ~ . + alcohol)
m3 <- update(m2, ~ . + residual.sugar)
m4 <- update(m3, ~ . + volatile.acidity)
m5 <- update(m4, ~ . + pH)
m6 <- update(m5, ~ . + chlorides)
mtable(m1, m2, m3, m4, m5, m6)

```

Based on our findings before, I have developed an approach with six models. 

Model 1: 
Density has a highly significant negative influence on the quality rating. It 
can explain about 9.4% of the variance of quality level ratings. The denser the 
wine, the worse the quality is rated.

Model 2: 
Alcohol has a highly significant positive influence on the quality rating. It 
can explain about 9.8% of the variance in quality level ratings. Its effect size
is less than the one of density, this can be explained by the different measure-
ment scales / the very small range of density variations, leading to very strong
p-value moves if density level is increased by 1. Interesting is that density
in model 2 shows a positive relation to quality, opposite to all other models.

The more alcohol, the better the wine quality is rated.

Model 3:
Residual sugar also has highly significant positive influence on the quality 
ratings. The more residual sugar, the better the wine quality is rated. The 
comprehensive regression model rejects the initial - bivariate correlation-based
- hypothesis that the effect of residual sugar on the quality is negativ, here 
is a positive relation.

Model 4: 
Volatile acidity has highly significant negative influence on the quality 
ratings. The higher the volatile acidity is, the worse the wine quality is 
rated. The comprehensive regression model rejects the initial - bivariate 
correlation-based - hypothesis that the effect of volatile acidity on the 
quality is negative, here is a positive relation.

Model 5:
pH value has a highly significant positive influence on the quality ratings. The
higher the pH value, the better the quality is perceived. 

Model 6:
I tested chlorides on top of model 5, but it was neither significant nor did it
improve the adjusted R-squared.

Adjusted R-squared: 
Model 1 and 2 contribute the most share of our overall adjusted R-squared, 
followed by volatile acidity and residual sugar. The overall value of 0.270
means that our model is able to explain about 27% of the variance in quality
ratings, which is quite good but also means that there are some 70% that are
not explained by the variables we built into our models.

```{r echo=FALSE, message=FALSE, warning=FALSE}

m1 <- lm(I(quality) ~ density, data = ww)
m2 <- update(m1, ~ . + alcohol)
m3 <- update(m2, ~ . + log10(residual.sugar))
m4 <- update(m3, ~ . + log10(volatile.acidity))
m5 <- update(m4, ~ . + pH)
m6 <- update(m5, ~ . + log10(chlorides))
mtable(m1, m2, m3, m4, m5, m6)

```

In this regression, I used some transformed variables. These transformations
make effect interpretations more difficult, therefore I will focus on the effect
direction and related changes in the adjusted R-squared that improve the overall 
model fit. 

Model 1: 
Density variable is the same, effects same as before.

Model 2: 
Alcohol variable is the same, effects same as before. 

Model 3:
Residual sugar is log10 transformed, effect direction the same but overall 
adjusted R-squared a little less than before. Yet we stick to the transformed
variable as the distribution requires transformation.

Model 4: 
Volatile acidity is log10 transformed, effect direction changed compared to
before: the higher volatile acidity, the lower the perceived wine quality. 
The adjusted R-squared slightly improved compared to the model before. We stick 
to the transformed variable as the distribution requires transformation.

Model 5:
pH variable is the same, effects a little less as before.  

Model 6:
Chlorides is log10 transformed, effect is negative and significant. The
adjusted R-squared now overall improved to 0.273. We stick to the transformed
variable as the distribution requires transformation.

Adjusted R-squared: 
With the model modifications, we have improved our adjusted R-squared from 0.270
to 0.273, a little improvement compared to before. 

Outliers:

Outlier values were not cut off for these variables to reduce complexity of the 
analysis, this data cleaning exercise should be thoroughly conducted with more 
time and is likely to improve the model performance. 

------

# Final Plots and Summary

### White wine quality plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Polishing needed (see above)

qplot(x = quality, data = ww, fill = I("brown")) +
  scale_x_continuous(breaks = c(3, 4, 5, 6, 7, 8, 9)) +
  xlab ('Perceived quality (3 - very poor / 9 - very high)') +
  ylab ('Number of wines in dataset') +
  ggtitle('White wine quality')

```

### White wine quality description

The perceived quality of white wine looks rather normally distributed. 

We can see here that the majority of white whines has a quality rating between
5 and 7, there are very limited counts at the outer boundaries of the histogram.

### Correlation matrix plot

```{r echo=FALSE, message=FALSE, warning=FALSE}

cor_mat2 <- cor(read.csv('wineQualityWhites.csv'))
corrplot(cor_mat2, method = "circle", title = "Correlation matrix", 
         mar=c(0,0,1,0))

```

### Correlation matrix description

We're focussing on quality, so here are some observations:
 - lower volatile acidity seems to have a slightly positive influence
 - same applies for chlorides and total sulfur dioxide
 - lower density seems to have a rather positive influence on quality
 - higher alcohol seems to have a strongly positive impact on quality 

Amongst the others, following observations are interesting: 

 - density has a strong negative correlation with alcohol and a strong positive
   relation with residual sugar, alcohol has a strong negative correlation with
   residual sugar. These are the strongest effects observed 
 - pH value is negatively related to fixed acidity
 - levels of free sulfur dioxide and total sulfur dioxide are positively 
   correlated with each other and positively correlated with density
 - fixed acidity is not related to volatile acidity
 - chlorides are positively correlated with density, but negatively with alcohol

We need to keep in mind the rule of thumb of small relatedness starting with 
r >= 0.3, which we can show here for some cases. Yet, this also depends on the 
case number, we have plenty of cases in our dataset.

### Relation density/alcohol/residual sugar plot

```{r echo=FALSE, message=FALSE, warning=FALSE}

p13 <- ggplot(aes(density, alcohol), data = ww) + 
  geom_point(alpha = 0.3, color = 'orange', size = 0.75) +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple') +
  xlab ('Density (g / cm^3)') +
  ylab ('Alc. in % of volume') +
  ggtitle('Density vs. alcohol (density outliers removed)')

p14 <- ggplot(aes(density, residual.sugar), data = ww) + 
  geom_point(alpha = 0.3, color = 'orange', size = 0.75) +
  xlim (quantile(ww$density, 0), quantile(ww$density, 0.95)) +
  ylim (quantile(ww$residual.sugar, 0), quantile(ww$residual.sugar, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple') +
  xlab ('Density (g / cm^3)') +
  ylab ('Res. sugar (g / dm^3)') +
  ggtitle('Density vs. residual sugar (density/sugar outliers removed)')

p15 <- ggplot(aes(alcohol, residual.sugar), data = ww) + 
  geom_point(alpha = 0.3, color = 'orange', size = 0.75) +
  ylim (quantile(ww$residual.sugar, 0), quantile(ww$residual.sugar, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple') +
  xlab ('Alcohol in % of volume') +
  ylab ('Res. sugar (g / dm^3)') +
  ggtitle('Alcohol vs. residual sugar (residual sugar outliers removed)')

grid.arrange(p13, p14, p15, ncol = 1)


```


### Relation density/alcohol/residual sugar description

Here, our suspicion of a relation between some of our independent variables is 
confirmed. We can observe the following based on scatter plots with a linear 
regression smoother (purple): 

 - Density increases with lower alcohol levels
 - Density increases with higher residual sugar levels
 - Alcohol decreases with higher residual sugar levels

The smoother lines confirm the correlation coefficients. This example also 
confirms that if we want to run a stellar regression analysis of what impacts
white wine quality, we need to be aware that the independent variables are 
related with one another. Without me being the wine expert, it might be the 
case that the level of residual sugar in a wine influences both density and 
alcohol and therefore indirectly the quality, even though sugar does not show
highly significant relations to wine quality in our correlation matrix.


### Regression Output

```{r echo=FALSE, message=FALSE, warning=FALSE}

m1 <- lm(I(quality) ~ density, data = ww)
m2 <- update(m1, ~ . + alcohol)
m3 <- update(m2, ~ . + log10(residual.sugar))
m4 <- update(m3, ~ . + log10(volatile.acidity))
m5 <- update(m4, ~ . + pH)
m6 <- update(m5, ~ . + log10(chlorides))
mtable(m1, m2, m3, m4, m5, m6)

```

In this regression, I used some transformed variables. These transformations
make effect interpretations more difficult, therefore I will focus on the effect
direction and related changes in the adjusted R-squared that improve the overall 
model fit. 

Model 1: 
Density has a highly significant negative influence on the quality rating. It 
can explain about 9.4% of the variance of quality level ratings. The denser the 
wine, the worse the quality is rated.

Model 2: 
Alcohol has a highly significant positive influence on the quality rating. It 
can explain about 9.8% of the variance in quality level ratings. Its effect size
is less than the one of density, this can be explained by the different measure-
ment scales / the very small range of density variations, leading to very strong
p-value moves if density level is increased by 1. Interesting is that density
in model 2 shows a positive relation to quality, opposite to all other models.

The more alcohol, the better the wine quality is rated. 

Model 3:
Residual sugar also has highly significant positive influence on the quality 
ratings. The more residual sugar, the better the wine quality is rated. The 
comprehensive regression model rejects the initial - bivariate correlation-based
- hypothesis that the effect of residual sugar on the quality is negativ, here 
is a positive relation.

Model 4: 
Volatile acidity has highly significant negative influence on the quality 
ratings. The higher the volatile acidity is, the worse the wine quality is 
rated. The comprehensive regression model rejects the initial - bivariate 
correlation-based - hypothesis that the effect of volatile acidity on the 
quality is negative, here is a positive relation.

Model 5:
pH value has a highly significant positive influence on the quality ratings. The
higher the pH value, the better the quality is perceived.  

Model 6:
Chlorides is log10 transformed, effect is negative and significant.

Adjusted R-squared: 
With the model modifications, we have improved our adjusted R-squared from 0.270
to 0.273, a little improvement compared to before. 

------

# Reflection

The univariate analysis uncovered insights about the distrubition of the data
that could be used for later work in the regression model. In theory, these 
might also have been applicable for the bivariate assessments, but would have
made the interpretation more difficult. As I wanted to focus the interpretation
on the regression models, I applied the transformations in the last stage of the
project. 

Bivariate and multivariate analyses led to insights that make it easier to 
understand which factors contribute to a high perceived white wine quality. 

Outlier values were not cut off for the transformed variables to reduce 
complexity of the analysis. This data cleaning exercise should be thoroughly 
conducted with more time and is likely to improve the model performance. 

Residual sugar in the log10 version has improved the model quality, but when
looking at the distribution of the variable, some other transformation might
lead to better effects.

As a next step, the approach and the results could be used for the red wine data
and shared with my local wine dealer to discuss the results. For wine producers, 
these insights can serve as key to bring their portfolio wines closer to the 
taste of the customers. 

A nice add-on to the dataset would be the price. A common hypothesis for wine 
goes: the more expensive, the higher the perceived quality. This has been dis-
proven in many lab-scale tests I've seen on TV, but would be interesting to see
how this relates in our sample of many thousand white and red wines. 
